---
title: 'Sampling and Fitting Priors: Binomial Example with Conjugate Priors'
author: "Ethan Alt"
date: "2023-02-01"
output: 
  html_document:
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Binomial Proportion ALC

* Consider an experiment that generates $y \sim \text{Binomial}\left(n, \theta\right)$. 

* We wish to determine the required value of $n$ such that average length of the 95\% credible interval is less than $w=0.10$ (ALC). \\

* We consider conjugate sampling and fitting priors as shown:

$$
\begin{eqnarray*}
    \pi_s(\theta) & \propto & \text{Beta}\left(\alpha_1 = 2.0,\alpha_2 = 8.0 \right) \\
    \pi_f(\theta) & \propto & \text{Beta}\left(\alpha_1 = 1.0,\alpha_2 = 1.0 \right)					
\end{eqnarray*}
$$

### R code

We begin by creating a function to compute ALC.

```{r, warning = FALSE, message = FALSE}
## Function to perform assurance calculation based on ALC for binomial proportion
#' @param w desired \math{1 - alpha / 2} credible interval width
#' @param alpha value for a \math{1 - alpha / 2} credible interval
#' @param sampling.shape1 first shape parameter for beta sampling prior
#' @param sampling.shape2 second shape parameter for beta sampling prior
#' @param fitting.shape1 first shape parameter for beta fitting prior
#' @param fitting.shape2 second shape parameter for beta fitting prior
#' @param sample.size vector of sample sizes to compute CI widths
#' @param ndatasets number of simulated data sets
assurance.alc <- function(
  w = 0.10, alpha = 0.05
  , sampling.shape1 = 2, sampling.shape2 = 8
  , fitting.shape1 = 1, fitting.shape2 = 1
  , sample.size = seq(150, 250, by = 1)
  , ndatasets = 10000
) {
  ## vector to store results
  widths <- numeric(length(sample.size))
  ## compute probabilities for quantiles based on alpha
  plow  <- alpha / 2
  phigh <- 1 - plow
  ## loop through sample sizes
  for ( i in seq_along(sample.size) ) {
    ## Obtain samples from sampling prior
    sprior.sample <- rbeta(
      n = ndatasets, shape1 = sampling.shape1, shape2 = sampling.shape2
    )
    ## Sample from predictive distribution of the sampling prior
    y.sprior.sample <- rbinom(n = ndatasets, size = sample.size[i], prob = sprior.sample)
    ## Compute parameters for posterior distribution (beta since conjugate)
    fitting.shape1.post <- y.sprior.sample + fitting.shape1
    fitting.shape2.post <- sample.size[i] - y.sprior.sample + fitting.shape2
    ## Compute average width of the CI
    widths[i] <-  mean(
        qbeta(phigh, fitting.shape1.post, fitting.shape2.post)
      - qbeta(plow, fitting.shape1.post, fitting.shape2.post)
    )
  }
  ## return results from simulation
  data.frame('n' = sample.size, 'width' = widths)
}
```

We may now call the created function and plot the results.

```{r, message = FALSE, warning = FALSE}
## Load packages
library(tidyverse)


## Get simulation
sims <- assurance.alc()

## Plot
ggplot(data = sims, aes(x = n, y = width)) + 
  geom_point() + 
  geom_line()
```

Notice that the curve is nonmonotone. This is due to Monte Carlo error.

For a better estimate, we can conduct scatterplot smoothing to obtain a
smooth estimate of the curve.

```{r}
fit.loess <- loess(width ~ n, data = sims)
sims$width.sm <- predict(fit.loess)
```

Now, we can obtain the estimated sample size

```{r}
(nopt <- sims$n[which(sims$width.sm < 0.10)[1]])
```

```{r}
ggplot(data = sims, aes(x = n)) +
  geom_point(aes(y = width), color = 'blue') + 
  geom_line(aes(y = width.sm), color = 'red') + 
  geom_hline(aes(yintercept = 0.10)) + 
  geom_vline(aes(xintercept = nopt)) +
  ylab("CI Width") + 
  xlab("Sample size") + 
  ggtitle("Assurance calculation: binomial proportion with ALC criterion")
```




