---
title: 'Sampling and Fitting Priors: Bayesian Power and Type I Error Rate'
author: "Ethan Alt"
date: "2023-02-02"
output: 
  html_document:
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Binomial Proportion Type I Error

* Consider an experiment that generates $y \sim \text{Binomial}\left(n, \theta\right)$. 

* We wish to determine the required value of $n$ such that the posterior probability $\Pr(\theta < 0.5 | D) \ge 0.975$

* We consider conjugate sampling and fitting priors.

### R code

We begin by creating a function to compute assurance by:

1. Sampling from a specified sampling prior and its predictive distribution

2. Analyzing the generated data from step (1) using the fitting prior.

3. Computing the resulting posterior probability

```{r, warning = FALSE, message = FALSE}
## Function to perform assurance calculation based on ALC for binomial proportion
#' @param value.null value of the null hypothesis
#' @param alpha value for a \math{1 - alpha / 2} credible interval
#' @param sampling.mean mean parameter for beta sampling prior
#' @param sampling.disp dispersion parameter for beta sampling prior. If infinity, a point mass at sampling.mean
#' @param fitting.mean mean parameter for beta fitting prior
#' @param fitting.disp dispersion shape parameter for beta fitting prior
#' @param sample.size vector of sample sizes to compute CI widths
#' @param ndatasets number of simulated data sets
assurance <- function(
  value.null
  , sample.size = seq(10, 100, by = 1)
  , alpha = 0.05
  , sampling.mean = 0.2, sampling.disp = 10
  , fitting.mean = 0.5, fitting.disp = 1
  , ndatasets = 10000
) {
  ## Compute shape parameters
  sampling.shape1 <- sampling.disp * sampling.mean
  sampling.shape2 <- sampling.disp * (1 - sampling.mean)
  fitting.shape1  <- fitting.disp * fitting.mean
  fitting.shape2  <- fitting.disp * (1 - fitting.mean)
  ## vector to store results
  post.probs <- matrix(nrow = ndatasets, ncol = length(sample.size))
  thresh     <- 1 - alpha/2
  ## loop through sample sizes
  for ( i in seq_along(sample.size) ) {
    ## Obtain samples from sampling prior
    if ( sampling.disp < Inf ) {
      sprior.sample <- rbeta(
        n = ndatasets, shape1 = sampling.shape1, shape2 = sampling.shape2
      )
      ## Sample from predictive distribution of the sampling prior
      y.sprior.sample <- rbinom(n = ndatasets, size = sample.size[i], prob = sprior.sample)
    } else {
      ## Sample from a fixed probability at sampling.mean
      y.sprior.sample <- rbinom(n = ndatasets, size = sample.size[i], prob = sampling.mean)
    }
    ## Compute parameters for posterior distribution (beta since conjugate)
    fitting.shape1.post <- y.sprior.sample + fitting.shape1
    fitting.shape2.post <- sample.size[i] - y.sprior.sample + fitting.shape2
    ## Compute posterior probability
    post.probs[, i] <- pbeta(value.null, fitting.shape1.post, fitting.shape2.post, lower.tail = TRUE)
  }
  data.frame(n = sample.size, assurance = colMeans(post.probs >= thresh))
}
```


### Bayesian power

We may now call the created function and plot the results.

```{r, message = FALSE, warning = FALSE}
## Load packages
library(tidyverse)
library(ggthemes)

## Power simulations
   ## Beta(2, 8) sampling prior
       ## Uniform fitting prior
       assurance.random.noninform <- assurance(
         value.null = 0.5, ndatasets = 100000
         , sampling.mean = 2/10, sampling.disp = 10
         , fitting.mean = 0.5, fitting.disp = 1
       )
       assurance.random.noninform$sampling.prior = 'Beta(2, 8)'
       assurance.random.noninform$fitting.prior = 'U(0, 1)'
       ## Beta(3, 6) fitting prior
       assurance.random.inform <- assurance(
         value.null = 0.5, ndatasets = 100000
         , sampling.mean = 2/10, sampling.disp = 10
         , fitting.mean = 2/10, fitting.disp = 10
        )
       assurance.random.inform$sampling.prior = 'Beta(2, 8)'
       assurance.random.inform$fitting.prior = 'Beta(2, 8)'
   ## Point mass prior at 0.2
       ## Uniform fitting prior
       assurance.pointmass.noninform <- assurance(
         value.null = 0.5, ndatasets = 100000
         , sampling.mean = 2/10, sampling.disp = Inf
         , fitting.mean = 0.5, fitting.disp = 1
       )
       assurance.pointmass.noninform$sampling.prior = 'Point mass'
       assurance.pointmass.noninform$fitting.prior = 'U(0, 1)'
       ## Beta(3, 6) fitting prior
       assurance.pointmass.inform <- assurance(
         value.null = 0.5, ndatasets = 100000
         , sampling.mean = 2/10, sampling.disp = Inf
         , fitting.mean = 2/10, fitting.disp = 10
        )
        assurance.pointmass.inform$sampling.prior = 'Point mass'
        assurance.pointmass.inform$fitting.prior = 'Beta(2, 8)'
```
```{r}
## Combine sims
sims <- rbind(assurance.random.noninform, assurance.random.inform, assurance.pointmass.noninform, assurance.pointmass.inform)

## Plot
ggplot(
  data = sims, 
  aes(x = n, y = assurance
      , color = fitting.prior, fill = fitting.prior
      , linetype = sampling.prior, shape = sampling.prior
  )
) + 
  geom_line(
    stat="smooth",method = "lm", formula = y ~ 0 + I(1/x) + I((x-1)/x)
    , alpha = 0.5, linewidth = 1
  ) +
  ggtitle("Bayesian power (assurance)") + 
  scale_color_tableau() + 
  scale_fill_tableau() + 
  geom_hline(yintercept = 0.8)
```


### Type I error

We now compute the Bayesian version of type I error using a point-mass 
sampling prior at the null hypothesis given by

$$
  \pi_s(\theta) = 1\{ \theta = 0.5 \}.
$$

```{r type1error}
## Type I error simulations using point mass prior at 0.5
## Uniform fitting prior
type1error.pointmass.noninform <- assurance(
 value.null = 0.5, ndatasets = 100000
 , sampling.mean = 0.5, sampling.disp = Inf
 , fitting.mean = 0.5, fitting.disp = 1
)
type1error.pointmass.noninform$sampling.prior = 'Point mass'
type1error.pointmass.noninform$fitting.prior = 'U(0, 1)'
## Beta(3, 6) fitting prior
type1error.pointmass.inform <- assurance(
 value.null = 0.5, ndatasets = 100000
 , sampling.mean = 0.5, sampling.disp = Inf
 , fitting.mean = 2/10, fitting.disp = 10
)
type1error.pointmass.inform$sampling.prior = 'Point mass'
type1error.pointmass.inform$fitting.prior = 'Beta(3, 6)'
```

```{r}
sims.type1error <- rbind(type1error.pointmass.noninform, type1error.pointmass.inform)

## Plot
ggplot(
  data = sims.type1error, 
  aes(x = n, y = assurance
      , color = fitting.prior, fill = fitting.prior
  )
) + 
  geom_line(
    stat="smooth",method = "lm", formula = y ~ 0 + I(1/x) + I((x-1)/x)
    , alpha = 0.5, linewidth = 1
  ) +
  ggtitle("Bayesian type I error") + 
  scale_color_tableau() + 
  scale_fill_tableau() +
  geom_hline(aes(yintercept = 0.025)) +
  ylab("Type I error")

```

