---
title: 'ESTEEM Study: Conjugate Prior for GLMs (CI Prior)'
author: "Ethan Alt"
date: "2023-01-26"
output: 
  html_document:
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
set.seed(123)
knitr::opts_chunk$set(echo = TRUE)
library(cmdstanr)
check_cmdstan_toolchain(fix = TRUE, quiet = TRUE)
register_knitr_engine(override = FALSE)
```

## Stan implementation of CI prior

We now write the Stan code for the CI prior. We assume a logistic regression 
model where we only possess summary statistics from the historical study.

We store the model in an object called `ci_model`.

```{cmdstan, output.var="ci_model", message=FALSE}
functions {
  real logistic_glm_conjugate_lpdf(vector beta, vector mu0, matrix X, real lambda) {
    int n = rows(X);
    vector[n] eta = X*beta;
    return lambda * sum( mu0 .* eta - log(1 + exp(eta)) );
  }
}
data {
  int<lower=0> n;                    // current data sample size
  int<lower=0> p;                    // number of covariates (excl. intercept and treatment effect)
  array[n] int<lower=0,upper=1> y;   // current data outcomes (binary)
  matrix[n,p] X;                     // design matrix of current data (incl intercept and trt)
  vector[p] betahat0;                // MLE of historical data set
  real<lower=0,upper=1> lambda;      // discounting parameter
}
transformed data {
  vector[n] mu0 = inv_logit(X * betahat0);
}
parameters {
  vector[p] beta;    // regression coefficients
}
model {
  // Conjugate prior
  target += logistic_glm_conjugate_lpdf(beta | mu0, X, lambda);
  // Likelihood
  target += bernoulli_logit_glm_lpmf(y | X, 0, beta);
}
```

## R code for analysis

### Obtaining the data

We can obtain the data directly from GitHub.

```{r data, message = FALSE, warning = FALSE}
library(tidyverse)
library(cmdstanr)
library(bayesplot)
library(posterior)
library(ggthemes)
library(kableExtra)
library(broom)


url <- 'https://raw.githubusercontent.com/ethan-alt/IntroBayesianAnalysis/main/Data'
esteem1 <- read.csv(file.path(url, 'esteem1.csv'))
esteem2 <- read.csv(file.path(url, 'esteem2.csv'))

## Keep only full data
esteem1 <- esteem1[complete.cases(esteem1), ]
esteem2 <- esteem2[complete.cases(esteem2), ]

## Declare placebo as reference group
esteem1$treatment <- factor(esteem1$treatment, levels = c('Placebo', '30 mg BID'))
esteem2$treatment <- factor(esteem2$treatment, levels = c('Placebo', '30 mg BID'))

## Create binary outcome
esteem1$pasi_resp <- with(esteem1, ifelse(pasi_pchg <= -75, 1, 0))
esteem2$pasi_resp <- with(esteem2, ifelse(pasi_pchg <= -75, 1, 0))

## Show table of current / historical data
with(esteem1, table(treatment, pasi_resp)) %>% kable() %>% 
  kable_classic(full_width = FALSE) %>%
  add_header_above(c(' ', 'Responders' = 2)) %>%
  pack_rows('Treatment', 1, 2)
with(esteem2, table(treatment, pasi_resp)) %>% kable() %>% 
  kable_classic(full_width = FALSE) %>%
  add_header_above(c(' ', 'Responders' = 2)) %>%
  pack_rows('Treatment', 1, 2)
```


### Constructing the Stan data

We now construct the Stan data.

```{r standata}
formula <- pasi_resp ~ treatment + scale(pasi_bsl) + scale(age) + I(scale(age)^2) + smoke + prior_sys

## Fit historical data MLE--pretend we don't observe historical data
hist.fit <- glm(formula, family = 'binomial', data = esteem1)
hist.fit %>% tidy() %>% kable(digits = 3) %>% kable_classic(full_width = FALSE)

## Get sample sizes --> compute lambda
(n0 <- nrow(esteem1))
(n  <- nrow(esteem2))
lambda <- 0.5 * min(n / n0, 1)

standata <- list(
  'y' = esteem2[, all.vars(formula)[1]]
  , 'X' = model.matrix(formula, esteem2)
  , 'betahat0' = coef(hist.fit)
  , 'lambda'   = lambda
)
standata$n  <- nrow(standata$X)
standata$n0 <- nrow(standata$X0)
standata$p  <- ncol(standata$X)
```


### Performing the MCMC sampling and conducting the analysis

We call the `sample()` method from `cmdstanr` to perform MCMC.

```{r mcmc, message = FALSE, warning = FALSE}
ci_smpl <- 
  ci_model$sample(
  data = standata, iter_warmup = 1000, iter_sampling = 2500, parallel_chains = 4
  , chains = 4, refresh = 0
)
```


```{r summary}
ci_smpl$summary(
  variables = NULL, mean, sd, 'q' = ~quantile2(.x, probs = c(0.025, 0.975))
) %>% kable(digits = 4) %>% kable_classic(full_width = FALSE)
```


### Comparison to a reference prior (no borrowing)

We can compare to an improper uniform prior by setting $\lambda = 0$.

#### R code for analysis with reference prior

```{r refprior}
ref_standata <- standata
ref_standata$lambda <- 0
ref_smpl <- suppressMessages(
  ci_model$sample(
    data = ref_standata, iter_warmup = 2000, iter_sampling = 2500
    , parallel_chains = 4, refresh = 0
  )
)
```
### Comparison between straPP and refernce prior

We plot the posterior densities of the treatment effect for each prior.

```{r}
draws.ci  <- ci_smpl$draws(format = 'data.frame')
draws.ref <- ref_smpl$draws(format = 'data.frame')
draws.ci$Prior  <- 'CI Prior'
draws.ref$Prior <- 'Reference'
plot.df <- rbind(draws.ci, draws.ref)

ggplot(
  plot.df, aes(x = .data[['beta[2]']], color = Prior, fill = Prior)
) +
  geom_density(alpha = 0.5) +
  scale_color_tableau() +
  scale_fill_tableau()

```



