---
title: 'ESTEEM Study: Propensity Score Integrated Power Prior (PSIPP)'
author: "Ethan Alt"
date: "2023-01-25"
output: 
  html_document:
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
set.seed(123)
knitr::opts_chunk$set(echo = TRUE)
library(cmdstanr)
check_cmdstan_toolchain(fix = TRUE, quiet = TRUE)
register_knitr_engine(override = FALSE)
options(knitr.table.format = "html")
```

## Stan implementation of psipp

We now write the Stan code for the PSIPP We assume a normal linear model where
we wish only to borrow from the historical controls.

We store the model in an object called `psipp_model`.

```{cmdstan, output.var="psipp_model", message=FALSE}
data {
  int<lower=0> n;                                 // current data sample size
  int<lower=0> n0;                                // historical data sample size
  int<lower=0> nstrata;                           // number of strata
  vector[n] y;                                    // current data outcomes
  vector[n0] y0;                                  // historical data outcomes
  array[n] int<lower=0,upper=1> trt;              // treatment assignment for current data
  array[n] int<lower=1,upper=nstrata> stratum;    // stratum assignment for current data
  array[n0] int<lower=1,upper=nstrata> stratum0;  // stratum assignment for historical data
  vector<lower=0,upper=1>[nstrata] a0;            // power prior parameters
  real mu_mean;                                   // mean hyperparameter for normal prior on mean
  real<lower=0> mu_sd;                            // SD hyperparameter for normal prior on mean
}
transformed data { 
  array[n] int trtindx;   // (1 = placebo; 2 = treatment)
  for ( i in 1:n ) {
    trtindx[i] = trt[i] + 1;
  }
}
parameters {
  matrix[nstrata, 2] mu;
  matrix<lower=0>[nstrata, 2] sigma;
}
model {
  // initial priors
  for ( j in 1:2 ) {
    mu[, j] ~ normal(mu_mean, mu_sd);
    sigma[, j] ~ cauchy(0, 1);
  }
  
  // historical data (power prior)
  for ( i in 1:n0 )
    target += a0[stratum0[i]] * normal_lpdf(y0[i] | mu[ stratum0[i], 1 ], sigma[ stratum0[i], 1 ]);  // only controls
  
  // likelihood
  for ( i in 1:n )
    target += normal_lpdf(y[i] | mu[ stratum[i], trtindx[i] ], sigma[ stratum[i], trtindx[i] ]);
}

generated quantities {
  vector[nstrata] trteff_stratum = mu[, 2] - mu[, 1];
  real trteff = mean(trteff_stratum);
}
```

## R code for analysis

### Obtaining the data

We can obtain the data directly from GitHub.

```{r data, message = FALSE, warning = FALSE}
library(tidyverse)
library(cmdstanr)
library(bayesplot)
library(posterior)
library(ggthemes)
library(kableExtra)
library(broom)


url <- 'https://raw.githubusercontent.com/ethan-alt/IntroBayesianAnalysis/main/Data'
esteem1 <- read.csv(file.path(url, 'esteem1.csv'))
esteem2 <- read.csv(file.path(url, 'esteem2.csv'))

## Keep only full data
esteem1 <- esteem1[complete.cases(esteem1), ]
esteem2 <- esteem2[complete.cases(esteem2), ]

## Subset for historical controls
esteem1.ctrls <- esteem1 %>% filter(treatment == 'Placebo')

## Declare placebo as reference group
esteem1$treatment <- factor(esteem1$treatment, levels = c('Placebo', '30 mg BID'))
esteem2$treatment <- factor(esteem2$treatment, levels = c('Placebo', '30 mg BID'))

## First few lines of esteem1
head(esteem1)

## First few lines of esteem2
head(esteem2)

## Construct pooled data set
pooled <- rbind(esteem1, esteem2) %>% mutate(studyid = factor(studyid))
```


### Constructing the Stan data

We now construct the Stan data.

#### Estimating the propensity scores

We use the full data to estimate the propensity scores

```{r ps}
## Set number of strata
nstrata <- 5

## Estimate propensity score (PS)
ps.formula <- studyid ~ scale(pasi_bsl) + scale(age) + I(scale(age)^2) + smoke + prior_sys
ps.glm     <- glm(ps.formula, family = 'binomial', data = pooled)
ps.glm %>% tidy() %>% kable(digits = 3) %>% kable_styling()

## Obtain fitted PS for all current data and historical controls
pooled.histctrls <- pooled %>%
  mutate(ps = predict(ps.glm, newdata = pooled, type = 'response')) %>% ## PS calculation
  filter(!( studyid == 1 & treatment == '30 mg BID' ))   ## drop treated historical
  
## Create strata based on propensity scores of CURRENT RCT
##   Note: this code simultaneously "trims" the PS--removes
##   external subjects whose PS is not in the range of the current.
breaks <- pooled.histctrls %>% 
  filter(studyid == 2) %>% 
  select(ps) %>% 
  unlist() %>%
  quantile(probs = seq(0, 1, 1 / nstrata))
breaks[1] <- 0
breaks[nstrata+1] <- 1

pooled.histctrls <- pooled.histctrls %>%
  mutate(stratum = cut(ps, breaks)) %>%   ## define strata. Will be NA if PS not in range of current data
  na.omit                                 ## trimming step

## Plot propensity scores by study ID
ggplot(pooled.histctrls, aes(x = ps, color = studyid, fill = studyid)) +
  geom_density(alpha = 0.5) +
  scale_color_tableau() + 
  scale_fill_tableau()

## Plot propensity scores by study ID and stratum
ggplot(pooled.histctrls, aes(x = ps, color = studyid, fill = studyid)) +
  geom_density(alpha = 0.5) +
  scale_color_tableau() + 
  scale_fill_tableau() +
  facet_wrap(~stratum)
```


#### Selection of $a_0$

We now determine the $a_{0m}$'s, where $m = 1, \ldots, 5$ indexes the 5 strata.

We first determine the nominal number of historical controls to borrow, 
$\tilde{n}_{00}$ we desire $\tilde{n_{00} + n_{10}} \approx n_{11}$,
where $n_{10}$ is the number of controls in the current data set and
$n_{11}$ is the number of treated in the current data set. Thus,
we should have
$$
  \tilde{n}_{00} = \min\{ n_{00}, n_{11} - n_{10} \}
$$

```{r nominal patients}
n11 <- pooled.histctrls %>% filter(studyid == 2, treatment == '30 mg BID') %>% nrow
n10 <- pooled.histctrls %>% filter(studyid == 2, treatment == 'Placebo') %>% nrow
n00 <- pooled.histctrls %>% filter(studyid == 1, treatment == 'Placebo') %>% nrow
ntilde00 <- min(n00, n11 - n10)
c(n11 = n11, n10 = n10, n00 = n00, ntilde00 = ntilde00)
```

To compute $a_0$, we first compute the PS overlap for each stratum.

```{r overlap}
library(overlapping)
(stratanames <- levels(pooled.histctrls$stratum))
ps.overlap <- sapply(
  stratanames, function(s) {
    ## create temporary data set for stratum
    df.stratum <- pooled.histctrls %>% 
      filter(stratum == s) %>% 
      select(studyid, ps)
    
    ## get list of propensity scores between current and historical
    pslist <- lapply(1:2, function(i) df.stratum$ps[df.stratum$studyid == i] )
    
    ## Compute overlap
    overlap(pslist, plot = FALSE)$OV
  }
)
names(ps.overlap) <- stratanames
round(ps.overlap, 3)

```

Next, we compute the nominal number of patients to borrow, given as

$$
  \lambda_{0m} = \min\left\{ \frac{r_m}{\sum_{l=1}^5 r_l} \tilde{n}_{00}, n_{00m} \right\},
$$

where $n_{00m} = $ number of observations among historical controls assigned to stratum $m$

```{r}
## Get number of historical controls assigned to each stratum (n00m)
nstratum.histctrls <- pooled.histctrls %>%
  filter(studyid == 1) %>%
  select(stratum) %>%
  table()

## Get number of current controls assigned ot each stratum (n1m)
nstratum.curctrls <- pooled.histctrls %>%
  filter(studyid == 2, treatment == 'Placebo') %>%
  select(stratum) %>%
  table()

## Get number of current trteated assigned ot each stratum (n1m)
nstratum.curtrts <- pooled.histctrls %>%
  filter(studyid == 2, treatment == '30 mg BID') %>%
  select(stratum) %>%
  table()

## Compute a0 based on ntilde00 and overlap
(w <- ps.overlap / sum(ps.overlap))                ## normalized overlap
lambda <- pmin( ntilde00 * w, nstratum.histctrls ) ## number of subjects to borrow in each stratum
sum(lambda)                                        ## total number of borrowed controls from historical data

cbind(
  'n_00m' = nstratum.histctrls,'lambda_0m' = lambda, 'n_10' = nstratum.curctrls
  , 'n_10star' = lambda + nstratum.curctrls, 'n11' = nstratum.curtrts
)
## Compute power prior parameter
(a0 <- lambda / as.numeric(nstratum.histctrls))

## Compute total number of controls, treated
c('ctrls' = n10 + sum(lambda), 'trt' = n11)
```
In this case, the total number of borrowed patients is `r sum(lambda)` and
the nominal number of patients to borrow is `r ntilde00``



### Constructing the Stan data

We now use the above to construct the Stan data.

```{r standata}
pooled.histctrls$treatment_n <- ifelse(pooled.histctrls$treatment == 'Placebo', 0, 1)
standata <- list(
  'y' = pooled.histctrls %>% filter(studyid == 2) %>% select(pasi_pchg) %>% unlist
  , 'y0' = pooled.histctrls %>% filter(studyid == 1) %>% select(pasi_pchg) %>% unlist
  , 'trt' = pooled.histctrls %>% filter(studyid == 2) %>% select(treatment_n) %>% unlist
  , 'stratum' = pooled.histctrls %>% filter(studyid == 2) %>% select(stratum) %>% unlist %>% as.numeric() 
  , 'stratum0' = pooled.histctrls %>% filter(studyid == 1) %>% select(stratum) %>% unlist %>% as.numeric()
  , 'a0'  = a0
  , 'mu_mean' = 0
  , 'mu_sd' = 10
  , 'nstrata' = nstrata
)
standata$n <- length(standata$y)
standata$n0 <- length(standata$y0)
```

Finally, we may call the MCMC sampler.

```{r mcmc}
psipp.smpl <- psipp_model$sample(
  data = standata, iter_warmup = 2000, iter_sampling = 2500, parallel_chains = 4
  , refresh = 0
)
```
We may summarize the output

```{r summarize}
psipp.smpl$summary(
  variables = c('trteff_stratum', 'trteff')
  , mean, sd, q = ~quantile2(.x, probs = c(0.025, 0.975))
)
```


### Comparison with reference prior

The code for the reference prior may be expressed as

```{cmdstan output.var="reference_model", message=FALSE}
data {
  int<lower=0> n;                                 // current data sample size
  vector[n] y;                                    // current data outcomes
  array[n] int<lower=0,upper=1> trt;              // treatment assignment for current data
  real mu_mean;                                   // mean hyperparameter for normal prior on mean
  real<lower=0> mu_sd;                            // SD hyperparameter for normal prior on mean
}
transformed data { 
  array[n] int trtindx;   // (1 = placebo; 2 = treatment)
  for ( i in 1:n ) {
    trtindx[i] = trt[i] + 1;
  }
}
parameters {
  vector[2] mu;
  vector<lower=0>[2] sigma;
}
model {
  // initial priors
  mu ~ normal(mu_mean, mu_sd);
  sigma ~ cauchy(0, 1);
  
  // likelihood
  for ( i in 1:n )
    target += normal_lpdf(y[i] | mu[trtindx[i]], sigma[trtindx[i]] );
}
generated quantities {
  real trteff = mu[2] - mu[1];
}
```


```{r reference}
ref.smpl <- reference_model$sample(
  data = standata, iter_warmup = 2000, iter_sampling = 2500, parallel_chains = 4
  , refresh = 0
)
ref.smpl$summary()
```

```{r compare_plot}
df.psipp <- cbind(prior = 'psipp', psipp.smpl$draws('trteff', format = 'data.frame') )
df.reference <- cbind(prior = 'reference', ref.smpl$draws('trteff', format = 'data.frame') )
plot.df <- rbind(df.psipp, df.reference)
ggplot(
  data = plot.df, aes(x = trteff, color = prior, fill = prior)
) +
  geom_density(alpha = 0.5) +
  scale_color_tableau() + 
  scale_fill_tableau()
```

