---
title: 'ESTEEM Study: Partial Borrowing Power Prior (PBPP)'
author: "Ethan Alt"
date: "2023-01-24"
output: html_document
---

```{r setup, include=FALSE}
set.seed(123)
knitr::opts_chunk$set(echo = TRUE)
library(cmdstanr)
check_cmdstan_toolchain(fix = TRUE, quiet = TRUE)
register_knitr_engine(override = FALSE)
```

## Stan implementation of PBPP

We now write the Stan code for the PBPP. We assume a normal linear model where
we wish only to borrow from the intercept and covariates.

We store the model in an object called `pbpp_model`.

```{cmdstan, output.var="pbpp_model", message=FALSE}
data {
  int<lower=0> n;            // current data sample size
  int<lower=0> n0;           // historical data sample size
  int<lower=0> p;            // number of covariates (incl. intercept and trteff)
  vector[n] y;               // current data outcomes
  vector[n0] y0;             // historical data outcomes
  matrix[n,p] X;             // design matrix of current data (incl intercept and trt)
  matrix[n0,p] X0;           // design matrix of historical data (incl intercept and trt)
  real<lower=0,upper=1> a0;  // discounting parameter
  real beta_mean;            // mean hyperparameter for initial prior on beta
  real<lower=0> beta_sd;     // sd hyperparameter for initial prior on beta
}
parameters {
  real intercept;         // intercept (borrowed)
  real trteff;            // treatment effect for current
  real trteff_hist;       // treatment effect for historical (not borrowed)
  vector[p-2] beta_x;     // covariates (borrowed)
  real<lower=0> sigma;    // SD parameter (borrowed)
}
transformed parameters {
  vector[p] beta = append_row(intercept, append_row(trteff, beta_x));           // = (intercept, trteff, beta_x)
  vector[p] beta_hist = append_row(intercept, append_row(trteff_hist, beta_x)); // = (intercept, trteff_hist, beta_x)
}
model {
  // Initial prior
  intercept ~ normal(beta_mean, beta_sd);
  trteff ~ normal(beta_mean, beta_sd);
  trteff_hist ~ normal(beta_mean, beta_sd);
  beta_x ~ normal(beta_mean, beta_sd);
  sigma ~ cauchy(0, 1);
  
  // Power prior (discounted likelihood)
  target += a0 * normal_id_glm_lpdf(y0 | X0, 0, beta_hist, sigma);
  
  // Likelihood
  y ~ normal_id_glm(X, 0, beta, sigma);
}
```

## R code for analysis

### Obtaining the data

We can obtain the data directly from GitHub.

```{r data, message = FALSE, warning = FALSE}
library(tidyverse)
library(cmdstanr)
library(bayesplot)
library(posterior)
library(ggthemes)


url <- 'https://raw.githubusercontent.com/ethan-alt/IntroBayesianAnalysis/main/Data'
esteem1 <- read.csv(file.path(url, 'esteem1.csv'))
esteem2 <- read.csv(file.path(url, 'esteem2.csv'))

## Keep only full data
esteem1 <- esteem1[complete.cases(esteem1), ]
esteem2 <- esteem2[complete.cases(esteem2), ]

## Declare placebo as reference group
esteem1$treatment <- factor(esteem1$treatment, levels = c('Placebo', '30 mg BID'))
esteem2$treatment <- factor(esteem2$treatment, levels = c('Placebo', '30 mg BID'))

## First few lines of esteem1
head(esteem1)

## First few lines of esteem2
head(esteem2)
```


### Constructing the Stan data

We now construct the Stan data.

```{r standata}
formula <- pasi_pchg ~ treatment + scale(pasi_bsl) + scale(age) + I(scale(age)^2) + smoke + prior_sys


standata <- list(
  'y' = esteem2[, all.vars(formula)[1]]
  , 'X' = model.matrix(formula, esteem2)
  , 'y0' = esteem1[, all.vars(formula)[1]]
  , 'X0' = model.matrix(formula, esteem1)
  , 'beta_mean' = 0
  , 'beta_sd' = 10
)
standata$n  <- nrow(standata$X)
standata$n0 <- nrow(standata$X0)
standata$p  <- ncol(standata$X)
(standata$a0 <- min(1, standata$n / standata$n0))
```


### Performing the MCMC sampling and conducting the analysis

We call the `sample()` method from `cmdstanr` to perform MCMC.

```{r mcmc, warning = FALSE, message = FALSE}
pbpp_smpl <- suppressMessages(
  pbpp_model$sample(
  data = standata, iter_warmup = 2000, iter_sampling = 4000, parallel_chains = 4
  , refresh = 0
)
)
pbpp_smpl$summary(
  variables = NULL, mean, sd, 'q' = ~quantile2(.x, probs = c(0.025, 0.975))
  , 'pr_lt_0' = ~mean(.x < 0)
)
```


```{r diagnostics}
pbpp_draws <- pbpp_smpl$draws(format = 'draws_df')
```


### Comparison to a reference prior (no borrowing)

We can fit a non-borrowing method by setting $a_0 = 0$

```{r reference}
ref_standata <- standata
ref_standata$a0 <- 0
ref_smpl <- suppressMessages(
    pbpp_model$sample(
    data = ref_standata, iter_warmup = 2000, iter_sampling = 4000
    , parallel_chains = 4, refresh = 0
  )
)
ref_smpl$summary(
  variables = NULL, mean, sd, 'q' = ~quantile2(.x, probs = c(0.025, 0.975))
  , 'pr_lt_0' = ~mean(.x < 0)
)
```

We can plot the posterior densities of the two treatment effects

```{r comparison}
trteff.pbpp <- pbpp_smpl$draws(variables = 'trteff', format = 'data.frame')
trteff.ref <- ref_smpl$draws(variables = 'trteff', format = 'data.frame')
trteff.pbpp$Prior <- 'PBPP (a0 = 0.48)'
trteff.ref$Prior <- 'Reference (a0 = 0)'
plot.df <- rbind(trteff.ref, trteff.pbpp)

ggplot(
  plot.df, aes(x = trteff, color = Prior, fill = Prior)
) +
  geom_density(alpha = 0.5) +
  scale_color_tableau() + 
  scale_fill_tableau()
```
