---
title: 'PLUTO: Power Prior Example'
author: "Ethan Alt"
date: "01/12/2023"
output: 
  html_document:
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(cmdstanr)
check_cmdstan_toolchain(fix = TRUE, quiet = TRUE)
register_knitr_engine(override = FALSE)
```


### Stan code
Below, we implement the Stan code for the power prior.
```{cmdstan, output.var="pp_model"}
data {
  int<lower=0> ntrt;                 // num. treated patients in current data set
  int<lower=0> nctrl;                // num. placebo patients in current data set
  int<lower=0,upper=ntrt> ytrt;      // num. responders in treated group
  int<lower=0,upper=nctrl> yctrl;    // num. responders in control group
  int<lower=0> J;                    // num. of historical data sets
  array[J] int<lower=0> ntrt0;       // num. treated patients in historical data sets
  array[J] int<lower=0> nctrl0;      // num. placebo patients in historical data sets
  array[J] int<lower=0> ytrt0;       // num. responders in treated group for historical data sets
  array[J] int<lower=0> yctrl0;      // num. responders in placebo group for historical data sets
  array[J] real<lower=0,upper=1> a0; // discounting parameter (could also be vector type)
  real<lower=0> theta_shape1;        // first shape hyperparameter for initial beta prior
  real<lower=0> theta_shape2;        // second shape hyperparameter for initial beta prior
}
parameters {
  real<lower=0,upper=1> theta1;  // probability of responding among treated
  real<lower=0,upper=1> theta0;  // probability of responding among placebo
}
model {
  // power priors
  for ( j in 1:J ) {
    target += a0[j] * binomial_lpmf(ytrt0[j] | ntrt0[j], theta1);    // power prior for treated in j-th data set
    target += a0[j] * binomial_lpmf(yctrl0[j] | nctrl0[j], theta0);  // power prior for placebo in j-th data set
  }
  theta1 ~ beta(theta_shape1, theta_shape2);   // initial prior
  theta0 ~ beta(theta_shape1, theta_shape2);   // initial prior
  
  // likelihood
  ytrt ~ binomial(ntrt, theta1);
  yctrl ~ binomial(nctrl, theta0);
}
generated quantities {
  real risk_difference = theta1 - theta0;
  real risk_ratio = theta1 * inv(theta0);  // = theta1 / theta0
  real odds_ratio = (theta1 * inv(1-theta1)) / (theta0 * inv(1-theta0));
}
```


In an R session, we would compile the Stan code as follows:
```{r, eval = FALSE}
library(cmdstanr)
pp_model <- cmdstan_model('path/to/file/filename.stan')
```


### Setup
First, we load required libraries
```{r, message = FALSE, warning = FALSE}
library(cmdstanr)
library(posterior)
library(bayesplot)
library(tidyverse)
library(ggthemes)
```

### Analysis in R
#### Creation of the Stan data
Now, we must tell `stan` what the variables named in the `data{}` block are.
This is done by creating a named list as follows.
```{r}
## Create data set
standata <- list(
    'ntrt'   = 53
  , 'nctrl'  = 39
  , 'ytrt'   = 28
  , 'yctrl'  = 17
  , 'J'      = 2
  , 'ntrt0'  = c(273, 290)
  , 'nctrl0' = c(275, 287)
  , 'ytrt0'  = c(118, 167)
  , 'yctrl0' = c(93, 125)
  , 'theta_shape1' = 0.5
  , 'theta_shape2' = 0.5
)
(n <- with(standata, ntrt + nctrl))
standata$a0 <- c(0.1, 0.1)

## Conduct MCMC sampling
pp_smpl <- pp_model$sample(
  data = standata, iter_warmup = 2000, iter_sampling = 2500
  , chains = 4, parallel_chains = 4, refresh = 0
)
```
To access the samples, we use the `draws` method, which in turn uses the
convenient `posterior` package. We will obtain the draws via a special `data.frame`
called a `draws_df`.
```{r}
pp_mcmc <- pp_smpl$draws(format = 'draws_df')
pp_mcmc %>% summarize_draws()
```

By default, the `summarize_draws()` method gives $90\%$ credible intervals. To
obtain a 95\% CI, we can do the following:
```{r}
pp_mcmc %>% summarize_draws(quantile2, .args = list(probs = c(0.025, 0.975)))
```

More generally, we can summarize different posterior quantities
```{r}
pp_mcmc %>% summarize_draws(
  'mean', 'sd', ~quantile(.x, probs = c(0.025, 0.975)), 'pgt0' = ~mean(.x > 0), 'pgt1' = ~mean(.x > 1)
)
```


In this case, the trial would be unsuccessful. Let's see what happens when
we change $a_{01} = a_{02} = 0.3$
```{r}
standata$a0 <- rep(0.2, 2)
pp_smpl2 <- pp_model$sample(
  data = standata, iter_warmup = 2000, iter_sampling = 2500
  , chains = 4, parallel_chains = 4, refresh = 0
)
pp_mcmc2 <- pp_smpl2$draws(format = 'draws_df')
pp_mcmc2 %>% summarize_draws(
  'mean', 'sd', ~quantile(.x, probs = c(0.025, 0.975)), 'pgt0' = ~mean(.x > 0), 'pgt1' = ~mean(.x > 1)
)
```


### Plots
The `bayesplot` package provides nice diagnostic plots. 


#### Trace plots

Trace plots show how the MCMC converges. Ideally, we want to see lines randomly
fluctuating.

```{r}
pp_mcmc2 %>% mcmc_trace()
```

#### Autocorrelation plots

Autocorrelation can sometimes be problematic as it can indicate that the
chain is "stuck". 

```{r}
pp_mcmc2 %>% mcmc_acf()
```

### Posterior plots

Oftentimes, it's useful to plot the posterior distribution of each parameter

```{r}
## Posterior histogram
pp_mcmc2 %>% mcmc_hist()
```



#### Comparing with reference prior

We set $\theta_1 \sim U(0, 1)$ and $\theta_0 \sim U(0, 1)$, only using the
PLUTO trial data.

```{cmdstan, output.var="refprior_mod", message = FALSE}
data {
  int<lower=0> ntrt;       // num. treated patients in historical data sets
  int<lower=0> nctrl;      // num. placebo patients in historical data sets
  int<lower=0> ytrt;       // num. responders in treated group for historical data sets
  int<lower=0> yctrl;      // num. responders in placebo group for historical data sets
}
parameters {
  real<lower=0,upper=1> theta1;     // response rate for treated
  real<lower=0,upper=1> theta0;     // response rate for placebo
}
model {
  ytrt ~ binomial(ntrt, theta1);
  yctrl ~ binomial(nctrl, theta0);
}
generated quantities {
  real risk_difference = theta1 - theta0;
  real risk_ratio = theta1 * inv(theta0);  // = theta1 / theta0
  real odds_ratio = (theta1 * inv(1-theta1)) * inv(theta0 * inv(1-theta0));
}
```


```{r refprior, message = FALSE, warning = FALSE}
refprior_smpl <- refprior_mod$sample(
  data = standata, chains = 4, parallel_chains = 4, iter_warmup = 2000
  , iter_sampling = 2500, refresh = 0
)
```


```{r plot}
df.refprior <- refprior_smpl$draws(variables = 'risk_difference', format = 'data.frame')
df.refprior$prior <- 'refprior'
df.pp  <- pp_smpl$draws(variables = 'risk_difference', format = 'data.frame')
df.pp$prior <- 'pp (a0 = 0.2)'
df.pp2 <- pp_smpl2$draws(variables = 'risk_difference', format = 'data.frame')
df.pp2$prior <- 'pp (a0 = 0.3)'
df.plot <- rbind(df.pp, df.pp2, df.refprior)
ggplot(
  df.plot, aes(x = risk_difference, fill = prior)
) + 
  geom_density(alpha = 0.5) + 
  scale_fill_tableau() + 
  ggtitle('Posterior densities')
```



