---
title: 'PLUTO: Power Prior Example'
author: "Ethan Alt"
date: "01/12/2023"
output: 
  html_document:
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(cmdstanr)
check_cmdstan_toolchain(fix = TRUE, quiet = TRUE)
register_knitr_engine(override = FALSE)
```


### Stan code
Below, we implement the Stan code for the power prior.
```{cmdstan, output.var="pbpp_model"}
data {
  int<lower=0> ntrt;                   // num. treated patients in current data set
  int<lower=0> nctrl;                  // num. placebo patients in current data set
  int<lower=0,upper=ntrt> ytrt;        // num. responders in treated group
  int<lower=0,upper=nctrl> yctrl;      // num. responders in control group
  int<lower=0> J;                      // num. of historical data sets
  array[J] int<lower=0> ntrt0;         // num. treated patients in historical data sets
  array[J] int<lower=0> nctrl0;        // num. placebo patients in historical data sets
  array[J] int<lower=0> ytrt0;         // num. responders in treated group for historical data sets
  array[J] int<lower=0> yctrl0;        // num. responders in placebo group for historical data sets
  array[J] real<lower=0,upper=1> a0;   // discounting parameter 
  real coef_mean;                      // mean hyperparameter for normal prior on regression coefficients
  real<lower=0> coef_sd;               // sd hyperparameter for normal prior on regression coefficients
}
parameters {
  real intercept;
  real trteff;
  real trteff_hist;
}
model {
  // power priors
  for ( j in 1:J ) {
    target += a0[j] * binomial_logit_lpmf(ytrt0[j] | ntrt0[j], intercept + trteff_hist); // power prior for treated in j-th data set
    target += a0[j] * binomial_logit_lpmf(yctrl0[j] | nctrl0[j], intercept);             // power prior for placebo in j-th data set
  }
  intercept ~ normal(coef_mean, coef_sd);
  trteff ~ normal(coef_mean, coef_sd);
  trteff_hist ~ normal(coef_mean, coef_sd);
  
  // likelihood
  ytrt ~ binomial_logit(ntrt, intercept + trteff);
  yctrl ~ binomial_logit(nctrl, intercept);
}
generated quantities {
  real theta1 = inv_logit(intercept + trteff);
  real theta0 = inv_logit(intercept);
  real risk_difference = theta1 - theta0;
  real risk_ratio = theta1 * inv(theta0);  // = theta1 / theta0
  real odds_ratio = (theta1 * inv(1-theta1)) / (theta0 * inv(1-theta0));
}
```

### Setup
First, we load required libraries
```{r, message = FALSE, warning = FALSE}
library(cmdstanr)
library(posterior)
library(bayesplot)
library(tidyverse)
library(ggthemes)
```

### Analysis in R

#### Creation of the Stan data

Now, we must tell `stan` what the variables named in the `data{}` block are.
This is done by creating a named list as follows.
```{r}
## Create data set
standata <- list(
    'ntrt'   = 53
  , 'nctrl'  = 39
  , 'ytrt'   = 28
  , 'yctrl'  = 17
  , 'J'      = 2
  , 'ntrt0'  = c(273, 290)
  , 'nctrl0' = c(275, 287)
  , 'ytrt0'  = c(118, 167)
  , 'yctrl0' = c(93, 125)
  , 'coef_mean' = 0
  , 'coef_sd'   = 10
)
(n <- with(standata, ntrt + nctrl))
standata$a0 <- c(0.2, 0.2)

## Conduct MCMC sampling
pbpp_smpl <- pbpp_model$sample(
  data = standata, iter_warmup = 2000, iter_sampling = 2500
  , chains = 4, parallel_chains = 4, refresh = 0
)
```

## Posterior summary
```{r}
pbpp_smpl$summary(
  variables = c('intercept', 'trteff', 'trteff_hist', 'risk_difference')
  , 'mean', 'sd', 'prob_gt_0' = ~mean(. > 0), 'q' = ~quantile2(., probs = c(0.025, 0.975))
)
```

### Plots
The `bayesplot` package provides nice diagnostic plots. 


#### Trace plots

```{r}
pbpp_smpl$draws() %>% mcmc_trace()
```


#### Comparing with reference prior

Here, we compare with the reference prior by setting $a_{0j} = 0$.


```{r refprior, message = FALSE, warning = FALSE}
refprior_standata <- standata
refprior_standata$a0 <- c(0, 0)
refprior_smpl <- pbpp_model$sample(
  data = refprior_standata, chains = 4, parallel_chains = 4, iter_warmup = 2000
  , iter_sampling = 2500, refresh = 0
)

refprior_smpl$summary(
  variables = c('intercept', 'trteff', 'trteff_hist', 'risk_difference')
  , 'mean', 'sd', 'prob_gt_0' = ~mean(. > 0), 'q' = ~quantile2(., probs = c(0.025, 0.975))
)
```


```{r plot}
df.refprior <- refprior_smpl$draws(variables = 'risk_difference', format = 'data.frame')
df.refprior$prior <- 'refprior'
df.pbpp  <- pbpp_smpl$draws(variables = 'risk_difference', format = 'data.frame')
df.pbpp$prior <- 'pbpp (a0 = 0.2)'
df.plot <- rbind(df.pbpp, df.refprior)
ggplot(
  df.plot, aes(x = risk_difference, fill = prior)
) + 
  geom_density(alpha = 0.5) + 
  scale_fill_tableau() + 
  ggtitle('Posterior densities')
```



