---
title: 'Pluto Trial: BHM'
author: "Ethan Alt"
date: "01/12/2023"
output: 
  html_document:
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
set.seed(123)
knitr::opts_chunk$set(echo = TRUE)
library(cmdstanr)
check_cmdstan_toolchain(fix = TRUE, quiet = TRUE)
register_knitr_engine(override = FALSE)
```


### Stan code
Below, we implement the Stan code for the power prior.
```{cmdstan, output.var="bhm_model", message=FALSE}
data {
  int<lower=0> J;                   // num. of data sets
  array[J] int<lower=0> ntrt;       // num. treated patients in historical data sets
  array[J] int<lower=0> nctrl;      // num. placebo patients in historical data sets
  array[J] int<lower=0> ytrt;       // num. responders in treated group for historical data sets
  array[J] int<lower=0> yctrl;      // num. responders in placebo group for historical data sets
  real betamean_mean;               // mean parameter for hierarchical mean
  real<lower=0> betamean_sd;        // sd parameter for hierarchical mean
}
parameters {
  vector[J] intercept;             // intercept
  vector[J] trteff;                // treatment effect
  real intercept_mean;             // hierarchical mean for intercept
  real trteff_mean;                // hierarchical mean for treatment effect
  real<lower=0> intercept_sd;      // hierarchical SD For intercept
  real<lower=0> trteff_sd;         // hierarchical SD for treatment effect
}
model {
  // Priors for hierarchical parameters
  intercept_mean ~ normal(betamean_mean, betamean_sd);
  trteff_mean ~ normal(betamean_mean, betamean_sd);
  intercept_sd ~ cauchy(0, 1);
  trteff_sd ~ cauchy(0, 1);
  // Loop through number of studies
  for ( j in 1:J ) {
    // Priors for model parameters
    target += normal_lpdf(intercept[j] | intercept_mean, intercept_sd);
    target += normal_lpdf(trteff[j] | trteff_mean, trteff_sd);
    // Likelihood contribution
    target += binomial_logit_lpmf(ytrt[j] | ntrt[j], intercept[j] + trteff[j]);
    target += binomial_logit_lpmf(yctrl[j] | nctrl[j], intercept[j]);
  }
}
generated quantities {
  real theta1 = inv_logit(intercept[1] + trteff[1]);
  real theta0 = inv_logit(intercept[1]);
  real risk_difference = theta1 - theta0;
  real risk_ratio = theta1 * inv(theta0);  // = theta1 / theta0
  real odds_ratio = (theta1 * inv(1-theta1)) * inv(theta0 * inv(1-theta0));
}
```



### Analysis in R
#### Load packages
```{r, message = F, warning = F}
library(cmdstanr)
library(posterior)
library(bayesplot)
library(tidyverse)
library(ggplot2)
library(ggthemes)
```

#### Creation of the Stan data

Now, we must tell `stan` what the variables named in the `data{}` block are.
This is done by creating a named list as follows.

```{r}
## Create data set
standata <- list(
    'J'      = 3
  , 'ntrt'  = c(53, 273, 290)
  , 'nctrl' = c(39, 275, 287)
  , 'ytrt'  = c(28, 118, 167)
  , 'yctrl' = c(17, 93, 125)
  , 'betamean_mean' = 0
  , 'betamean_sd' = 5
)

## Conduct MCMC sampling
bhm_smpl <- bhm_model$sample(
  data = standata, iter_warmup = 2000, iter_sampling = 2500
  , chains = 4, parallel_chains = 4, refresh = 0
)
```

We can obtain posterior mean, SD, 95%, and posterior probabilities as follows:
```{r}
parmnames <- c('intercept', 'trteff')
suffix    <- c('[1]', '_mean', '_sd')
pars      <- apply(expand.grid(parmnames, suffix), 1, paste, collapse="")
pars      <- c(pars, 'theta1', 'theta0', 'odds_ratio', 'risk_difference', 'risk_ratio')

bhm_smpl$summary(
  variables = pars,'mean', 'sd', 'q' = ~quantile2(.x, probs = c(0.025, 0.975))
  , 'p_gt_0' = ~mean(.x > 0), 'p_gt_1' =~ mean(.x > 1)
)
```

The trace plots are:

```{r}
bhm_mcmc <- bhm_smpl$draws()
bhm_mcmc %>% mcmc_trace()
```


Plotting the densities, by chain

```{r}
bhm_mcmc %>% mcmc_dens_overlay()
```

and the histogram combining chains...

```{r}
bhm_mcmc %>% mcmc_hist()
```


#### Comparing with a reference prior

We set $\theta_1 \sim U(0, 1)$ and $\theta_0 \sim U(0, 1)$, only using the
PLUTO trial data.

```{cmdstan, output.var="refprior_mod", message = FALSE}
data {
  int<lower=0> ntrt;       // num. treated patients in historical data sets
  int<lower=0> nctrl;      // num. placebo patients in historical data sets
  int<lower=0> ytrt;       // num. responders in treated group for historical data sets
  int<lower=0> yctrl;      // num. responders in placebo group for historical data sets
}
parameters {
  real<lower=0,upper=1> theta1;     // response rate for treated
  real<lower=0,upper=1> theta0;     // response rate for placebo
}
model {
  ytrt ~ binomial(ntrt, theta1);
  yctrl ~ binomial(nctrl, theta0);
}
generated quantities {
  real risk_difference = theta1 - theta0;
  real risk_ratio = theta1 * inv(theta0);  // = theta1 / theta0
  real odds_ratio = (theta1 * inv(1-theta1)) * inv(theta0 * inv(1-theta0));
}
```


```{r refprior, message = FALSE, warning = FALSE}
standata_ref <- standata
vars <- c('ytrt', 'yctrl', 'ntrt', 'nctrl')
for ( x in vars )
  standata_ref[[x]] <- standata_ref[[x]][1]
refprior_smpl <- refprior_mod$sample(
  data = standata_ref, chains = 4, parallel_chains = 4, iter_warmup = 2000
  , iter_sampling = 2500, refresh = 0
)
```

```{r plot}
df.refprior <- refprior_smpl$draws(variables = 'risk_difference', format = 'data.frame')
df.refprior$prior <- 'refprior'
df.bhm <- bhm_smpl$draws(variables = 'risk_difference', format = 'data.frame')
df.bhm$prior <- 'bhm'
df.plot <- rbind(df.bhm, df.refprior)
ggplot(
  df.plot, aes(x = risk_difference, fill = prior)
) + 
  geom_density(alpha = 0.5) + 
  scale_fill_tableau() +
  ggtitle('Posterior densities')
```
