---
title: 'Single-Arm Sequential Monitoring Example'
author: "Ethan Alt"
date: "2023-02-02"
output: 
  html_document:
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(kableExtra)
```

## R Code for Sequential Monitoring

The following function `SA_SEQ` computes summary statistics on the number of interim analyses, the sample size, the probabilities of stopping for efficacy/futility, etc., for a single-arm Bayesian sequential trial with a prespecified handicap prior.

```{r}
#' Compute sample size with Bayesian monitoring
#' Function that computes the required sample size with Bayesian monitoring
#' for a single-arm trial with a binary outcome.
#' 
#' @param nSim: number of simulations
#' @param pi0: null response threshold
#' @param piActual: actual value of response rate
#' @param nMax: maximum sample size to evaluate
#' @param delta: number of observed outcomes between interim analyses
#' @param h: proportion of maximum sample size that results in effective sample size
#' @param alpha: compare efficacy/futility probabilities to alpha/2
#' @param seed: integer for the random number generator
#' 
#' @return 
SA_SEQ <- function(nSim, pi0, piActual, nMax, delta, h, alpha, seed = 1){
  
  # Set seed and save values used for all iterations
  set.seed(seed)
  K0 <- ceiling(nMax/delta)               # max number of analyses
  alpha0 <- pi0 * K0 * delta * h          # first shape parameter in prior
  gamma0 <- (1 - pi0) * K0 * delta * h    # second shape parameter in prior
  
  # Matrix to store results
  results <- matrix(0, nrow = nSim, ncol = 10)
  
  ## Create progress bar
  pb <- txtProgressBar(
    min = 0, max = nSim, style = 3,width = 50,char = "="
  )   
  
  ## Loop through number of simulations
  for(sim in 1:nSim) {
    setTxtProgressBar(pb, sim)
    
    y <- 0
    n <- 0
    for(k in 1:K0) {
      # Sample delta additional observations
      y <- y + rbinom(1, delta, piActual)
      n <- n + delta
      
      # Calculate probabilities of efficacy and futility (posterior probabilities of observed data)
      pp_Eff <- pbeta(pi0, y + alpha0, (n - y) + gamma0, lower.tail = FALSE)
      pp_Fut <- pbeta(pi0, y + alpha0, (n - y) + gamma0, lower.tail = TRUE)
      
      # Compare efficacy and futility probabilities to prespecified thresholds
      futility <- pp_Eff < (alpha/2)
      efficacy <- pp_Fut < (alpha/2)
      
      # Calculate the mean, lower bound, and upper bound of credible interval
      mean.val <- (y + alpha0) / (n + alpha0 + gamma0)
      lower.val <- qbeta(0.025, y + alpha0, (n - y) + gamma0)
      upper.val <- qbeta(0.975, y + alpha0, (n - y) + gamma0)
      
      # Check if trial should stop for efficacy/futility or if last analysis
      if(futility + efficacy == 1 | k == K0){
        results[sim,] <- c(k, n, mean.val, lower.val, upper.val, futility, efficacy,
                           futility + efficacy, pp_Eff, pp_Fut)
        break
      }
    }
  }
  
  # Save summary statistics in a data frame
  var.names <- c("k", "n", "mean", "lower", "upper", "futility", "efficacy",
                 "error", "pp_Eff", "pp_Fut")
  lab.names <- c("Number of Interim Analyses", "Sample Size", "Sample Proportion",
                 "Lower Limit of 95% CI", "Upper Limit of 95% CI", "Futility", "Efficacy",
                 "Futility or Efficacy", "Prob(Efficacy)", "Prob(Futility)")
  summary.df <- data.frame( Variable = var.names,
                            Label = lab.names,
                            Mean = round(colMeans(results), 3),
                            StdDev = round(apply(results, 2, sd), 3),
                            Median = round(apply(results, 2, median), 3),
                            Minimum = round(apply(results, 2, min), 3),
                            Maximum = round(apply(results, 2, max), 3) )
  close(pb)
  return(summary.df)
  
}
```

#### Case when actual value of response rate is $\pi = 0.3$

```{r, results = "hide"}
# Case when actual value of response rate is pi = 0.3
results.pi.3 <- SA_SEQ(nSim = 500000, pi0 = 0.3, piActual = 0.3, nMax = 90,
                       delta = 15, h = 0.3, alpha = 0.05, seed = 1)
```
```{r}
results.pi.3 %>%
  kable(digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```


#### Case when actual value of response rate is $\pi = 0.5$
```{r, results = "hide"}
# Case when actual value of response rate is pi = 0.5
results.pi.5 <- SA_SEQ(nSim = 500000, pi0 = 0.3, piActual = 0.5, nMax = 90,
                       delta = 15, h = 0.3, alpha = 0.05, seed = 1)
```
```{r}
results.pi.5 %>%
  kable(digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

